\documentclass[a4paper, 11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fix-cm}
\usepackage{textcomp}
\usepackage[a4paper, margin=3cm]{geometry}
\usepackage[titletoc, title]{appendix}
\usepackage{booktabs}   % for \toprule, \midrule, \bottomrule
\usepackage{makecell}   % for \makecell
\usepackage{graphicx}   % for \resizebox
\usepackage{array}      % for >{\raggedright\arraybackslash}p{} column type
\usepackage{float} 
\usepackage{listings}

\usepackage{color}
\usepackage{booktabs}
\usepackage[all]{nowidow}
\usepackage[dvipsnames]{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{acronym}
\usepackage{graphicx}
\usepackage{url}
\usepackage{titlesec}
\usepackage{csquotes}
\usepackage{enumitem}
% math and algorithms
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
% tables
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{verbatimbox}
\usepackage{verbatim}

% general formatting
\usepackage{graphicx}

% xml and code listings
\usepackage{listings}
\usepackage{xcolor}

\lstdefinelanguage{XML}{
  morestring=[b]",
  morecomment=[s]{<!--}{-->},
  stringstyle=\color{teal},
  identifierstyle=\color{black},
  keywordstyle=\color{blue},
  morekeywords={xmlns,version,type,id,name,value}
}

\lstset{
  language=XML,
  basicstyle=\ttfamily\small,
  numbers=left,
  numberstyle=\tiny\color{gray},
  frame=single,
  breaklines=true,
  showstringspaces=false,
  tabsize=2,
  captionpos=b
}

\lstdefinelanguage{YAML}{
    keywords={true, false, null, !!str, !!map, !!seq},
    keywordstyle=\color{blue}\bfseries,
    basicstyle=\ttfamily\footnotesize,
    comment=[l]{\#},
    commentstyle=\color{gray}\itshape,
    stringstyle=\color{red},
    moredelim=[l]{\&},
    moredelim=[l]{*},
    moredelim=**[il][\color{green}]{:},
    morestring=[b]',
    morestring=[b]"
}

\lstdefinelanguage{C++}{
    language=C++,
    morekeywords={
        alignas, alignof, and, and_eq, asm, atomic_cancel, atomic_commit, atomic_noexcept,
        auto, bitand, bitor, break, case, catch, class, compl, concept, consteval, constexpr,
        constinit, const_cast, continue, co_await, co_return, co_yield, decltype, default,
        delete, do, dynamic_cast, else, enum, explicit, export, extern, false, for, friend,
        goto, if, import, inline, module, mutable, namespace, new, noexcept, not, not_eq,
        nullptr, operator, or, or_eq, private, protected, public, reflexpr, register,
        reinterpret_cast, requires, return, sizeof, static, static_assert, static_cast,
        struct, switch, synchronized, template, this, thread_local, throw, true, try, typedef,
        typeid, typename, union, using, virtual, volatile, while, xor, xor_eq
    },
    sensitive=true,
    morecomment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]",
    morestring=[b]',
    moredelim=**[is][\color{blue}]{@}{@},
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{gray}\itshape,
    stringstyle=\color{red},
    basicstyle=\ttfamily\footnotesize,
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    numbersep=10pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    frame=single,
    tabsize=4,
    captionpos=b,
    breaklines=true,
    breakatwhitespace=true,
    escapeinside={\%*}{*)}
}

\usepackage{transparent}
\usepackage{eso-pic}
\usepackage[section]{placeins}
\usepackage{setspace}
\usepackage{parskip}

\usepackage{subcaption}

\renewcommand\thefigure{%
\thesection.\arabic{figure}}
\renewcommand\thesubfigure{%
\thesection.\arabic{figure}.\arabic{subfigure}}
\renewcommand\thetable{%
\thesection.\arabic{table}}

\usepackage[main=english, ngerman]{babel}

% we use the cleveref package to refer to figures, sections, etc.
% instead of "Figure~\ref{fig:example}", write only "\cref{fig:example}" and the word "Figure" (or table, etc) will be inserted normally
\usepackage[noabbrev,capitalise]{cleveref}

\usepackage[
    maxbibnames=99,
    style=numeric,
    url=false,
    backend=bibtex8,
    sortcites=true,
]{biblatex}
\bibliography{refs}
\DeclareFieldFormat[online]{urldate}{Last accessed: #1}
\DeclareFieldFormat{eprint}{arXiv: \href{https://arxiv.org/abs/#1}{#1}}
\DeclareFieldFormat[report]{title}{``#1''}


\newcommand{\projectTitle}{Energy-aware Co-location of Scientific Workflow Tasks}
\newcommand{\thesisType}{Master's Thesis}
\newcommand{\authors}{Niklas Fomin}
\newcommand{\matrikel}{464308}
\newcommand{\authorEmail}{\href{mailto:niklas.fomin@campus.tu-berlin.de}{niklas.fomin@campus.tu-berlin.de}}
\newcommand{\examinera}{Prof.~Dr.~habil.~Odej Kao}
\newcommand{\examinerb}{Prof.~Dr. Volker Markl}
\newcommand{\supervisor}{Jonathan Bader}

\newcommand{\projectYear}{2025}
\newcommand{\facultyName}{Fakultät Elektrotechnik und Informatik}
\newcommand{\departmentName}{Fachgebiet Distributed and Operating Systems}
\newcommand{\myparagraph}[1]{\paragraph{#1}\mbox{}\\}

\begin{document}
\input{front.tex}
\clearpage
\pagestyle{plain}

\newpage

\section*{KI}
(https://www.chicagomanualofstyle.org/qanda/data/faq/topics/Documentation/faq0422.html)
If you need a more formal citation—for example, for a student paper or for a research article—a numbered footnote or endnote might look like this:

1. Text generated by ChatGPT, OpenAI, March 7, 2023, https://chat.openai.com/chat.

If you've edited the AI-generated text, you should say so in the text or at the end of the note (e.g., "edited for style and content").

If the prompt hasnt been included in the text, it can be included in the note:

1. ChatGPT, response to "Explain how to make pizza dough from common household ingredients," OpenAI, March 7, 2023.


\newpage

\section*{Acknowledgements}
I thank all of them for the support....



\newpage
\section*{Abstract}


Most applications running on supercomputers achieve only a fraction of the peak performance of the system. In this paper we analyze the performance and energy efficiency of co-scheduling one memory bandwidth bound and one compute bound application on the same node. We present auto pin+, a tool designed to monitor and optimize co-scheduling of applications. Our analysis shows that co-scheduling can improve both energy efficiency and overall throughput of a supercomputer. At best, runtime can be decreased by 28\% and the energy consumption by 12\%, respectively, compared to best case dedicated execution. The overall efficiency however strongly depends on the ratio of jobs available in the queue. We furthermore present a simple adaptive strategy depending on the available jobs in the queue.

The combined contribution of processing units and memory devices to a data center's total energy consumption typically ranges between 70\% and 80\%. When scientific workflows are executed in HPC data centers and due to parallelized execution their tasks are co-located—either on the same physical node or within the same virtual container—they must share these underlying compute and memory resources. A high level of resource sharings slows down the processing duration and increases the energy consumption compared to when executing them exclusively since they dont interfer with each other. Therefore, if possible workflow tasks with a high shared resource use should be allocated to compute resources to avoid contention, thus avoiding slowdown. In this thesis we present \textit{ShaReComp}, a multi-staged approach that characterizes tasks in terms of temporal signatures and applies dissimilarity-based clustering to group tasks with complementary resource usage. We integrate \textit{ShaReComp} into simulated workflow execution where we implement 4 co-location aware scheduling algorithms alongside blueprint baselines, that we call  textit{ShaRiff}. We evaluate our approach using 9 real-world scientific workflows, and compare the effects of random co-location against our approaches. Our results show that \textit{ShaReComp} does this and \textit{ShaRiff} can reduce the overall energy consumption by up to 18\% and the makespan by up to 11\% compared to random co-location.

% GERMAN
\clearpage
\begin{otherlanguage}{ngerman}
    \section*{Kurzfassung}
    FaaS ist ein innovatives neues Servicemodell, das sich mit dem aktuellen Vormarsch des Cloud Computing entwickelt hat.
    Softwareentwicklende können ihre Anwendungen schneller und mit der erforderlichen Flexibilität und Skalierbarkeit bereitstellen und gleichzeitig den Wartungsaufwand für die Infrastruktur sehr gering halten.
    Diese Vorteile sind im Edge-Computing sehr wünschenswert, da sich ständig ändernde Technologien und Anforderungen schnell umgesetzt werden müssen und die Fluktuation und Heterogenität der Service-Consumer ein wichtiger Faktor ist.
    Edge-Nodes können jedoch häufig nur einen Bruchteil der Leistung von Cloud-Computing-Infrastruktur bereitstellen, was die Ausführung herkömmlicher FaaS-Plattformen unmöglich macht.
    In dieser Arbeit stellen wir einen neuen Ansatz für FaaS eine Platform vor, die von Grund auf unter Berücksichtigung von Edge-Computing- und IoT-Anforderungen entwickelt wurde.
    Um den Overhead so gering wie möglich zu halten, nutzen wir CoAP als Messaging-Protokoll und Docker, um Applikationen voneinander zu isolieren, während Container wiederverwendet werden, um die bestmögliche Leistung zu erzielen.
    Wir präsentieren auch eine Proof-of-Concept-Implementierung unseres Designs, die wir mit einem eigenen Benchmarking-Tool getestet haben, und vergleichen unsere Ergebnisse mit Benchmarks von Lean OpenWhisk, einer weiteren FaaS-Plattform für die Edge.
    Wir stellen fest, dass unsere Plattform Lean OpenWhisk in Bezug auf Latenz und Durchsatz in allen Tests übertreffen kann, Lean OpenWhisk jedoch höhere Erfolgsraten bei einer geringen Anzahl gleichzeitiger Clients aufweist.
\end{otherlanguage}

\clearpage

\tableofcontents
\clearpage

\include{chapters/01_introduction}
\clearpage
\include{chapters/02_background}
\clearpage
\include{chapters/03_relatedwork}
\clearpage
\include{chapters/04_approach}
\clearpage
\include{chapters/05_implementation}
\clearpage
\include{chapters/06_evaluation}
\clearpage

\include{chapters/07_discussion}
\clearpage

\include{chapters/08_conclusionfuturework}
\clearpage

\include{chapters/09_appendix}

\newpage
\printbibliography
\end{document}